---
title: "Capstone Difference in Difference Report"
author: "Cindy Elder, Max Bi and Gabby Lopez"
date: "`r Sys.Date()`"
output: word_document
---

# Old version with rates per appointment, not per patient

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(zoo) # to extract month out of the standard date format
library(rstudioapi) # to allow us to use file.path 
library(did) # package from other professors who require time change in start
library(geepack) #geeglm
library(gtsummary)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
# Do we have to run the models in here to get the printouts? yes


# DID Models
#rm(list = ls())
# Import dataframes DataCleaningDID.R script


fp_visits = file.path(getwd(), "Analysis Data/all_visit_types.csv")
all_visit_types <- read.csv(fp_visits)
unique(all_visit_types$year)

#all_visit_types 
# Already marked Visit Types with Marshallese and NHW
# all_visit_types  has M indicator from previous DataCleaningDID.R script, which shows 
# 1 = marshallese, 0 = Non-Hispanic White, NA = Other not in our analysis)


# start building dataframe
# Change to year

did_visit_types_year <- all_visit_types %>% filter(Date >= "2018-08-31", Date <= "2019-08-31"  | # pretreat
                                                Date >= "2021-08-31", Date <= "2022-08-31") # post treat

# start_date = "2018-08-31", end_date = "2019-08-31") # All service lines = total population
# 
# start_date = "2021-08-31", end_date = "2022-08-31") # All service lines = total population


# remove the NA, not our target pop
did_visit_types_year <- did_visit_types_year %>% filter(!is.na(marsh)) 
# removed 1/3 of visits not our target pop



did_visit_types_year <- did_visit_types_year %>% mutate(ER = if_else( ((ServiceLine == "Emergency") &
                                                                    (Date >= "2019-06-01" & Date <= "2019-08-31" |
                                                                       Date >= "2022-06-01" & Date <= "2022-08-31" ) ), 1, 0),
                                              PCP = if_else( ( ( ServiceLine == "Primary Care") &
                                                                 (Date >= "2019-06-01" & Date <= "2019-08-31" |
                                                                    Date >= "2022-06-01"& Date <= "2022-08-31" ) ), 1, 0))



######################################################
########## BUILD DATA FRAME FOR DID ##################
######################################################




# size of our populations in this version of the study

# 
Marshallese_yr <- did_visit_types_year %>% filter(marsh == 1) 
length(unique(Marshallese_yr$UniqueIdentifier))
# 476 Marshallese visited these 2 clinics this year
# gabby got 495

NHW_yr <- did_visit_types_year %>% filter(marsh == 0) 
length(unique(NHW_yr$UniqueIdentifier))
# 18586 NonHispanic Whites visited these 2 clinics this year
# gabby got 20177


# we will have some of the same patients who appear in both time frames, but more in the 2nd time period

# and all Non-Hispanic White uniqueID for any type of appointment per year



# number of ER visits per month/ Population size per year

# numerator from the  ER2016_2025_yearmonth_marsh and 

# add months 
#all_visit_types <- all_visit_types %>% mutate(yearmonth = as.Date(zoo::as.yearmon(all_visit_types$Date, "%m/%d/%Y")))

# denominator from pop_size_year 




# Jan 2017 to Aug 2019 ER Pretrends (about 2.5 years)

# Sept 2019 to Sept 2022 DID ER Estimates for CHW 1 and 2

# Oct 2022 to Dec 2024 additional ER trends from CHW 3 and 4, KOH and Change to Control Pop Insurance (Medicaid Unwinding ~ Dec 2022) 


# ! Note that ER visits are recorded from 2017 forward. 

# But primary care provider (PCP) visits are only recorded from 2018 forward
# Jan 2018 to Aug 2019 PCP Pretrends (about 1.5 years)

# Sept 2019 to Sept 2022 DID PCP Estimates for CHW 1 and 2

# Oct 2022 to Dec 2024 additional PCP trends from CHW 3 and 4, KOH and Change to Control Pop Insurance (Medicaid Unwinding ~ Dec 2022) 



# How to we account for correlation with longitutindal data  
# using unbalanced panel in the DID package




################### GEEGLM #################
############################################

# Standford said they use GEEGLM for correlation 
# https://diff.healthpolicydatascience.org/


# Use GEEGLM for all



# Center the data so the intercept has better interpretation- Won't effect Slope or p-values

did_visit_types_year <- did_visit_types_year %>% mutate(year_center = (year - 2019) )




# I think we need to change the format of the data. 
# if we have the did_visit_types_year for the geeglm I think it will give us the number of expected ER visits per visit! 
# not per patient. 
# If we make a table with uniqueID, ER, marsh and year_center I think we get the data we need

# PCP_per_person <- as.data.frame(with(did_visit_types_year, table(UniqueIdentifier, marsh, PCP, year_center)))


did_visit_types_year <- did_visit_types_year %>% mutate(pre_year = if_else((year == 2018 | year == 2019), 1, 0),
                                                        post_year = if_else((year == 2021 | year == 2022), 1, 0) )
                                                                           

Marshallese <- did_visit_types_year %>% filter(marsh == 1) 
#colnames(Marshallese)
MarshalleseUniqueID <- unique(Marshallese$UniqueIdentifier)

#number of unique Marshallese
length(unique(MarshalleseUniqueID))
# 476
# gabby got 495

Control <-did_visit_types_year %>% filter(marsh == 0) 
ControlUniqueID <- unique(Control$UniqueIdentifier)

# Number of unique controls 
length(unique(ControlUniqueID))
# 18586
# gabby got 20177


PCP_per_person <- did_visit_types_year %>% group_by(UniqueIdentifier, post_year) %>% summarize( sum(PCP))
PCP_per_person <- PCP_per_person %>%  mutate(marsh = if_else((UniqueIdentifier %in% MarshalleseUniqueID), 1,
                                                             if_else((UniqueIdentifier %in% ControlUniqueID), 0, NA) ) )


ER_per_person <- did_visit_types_year %>% group_by(UniqueIdentifier, post_year) %>% summarize( sum(ER))
ER_per_person <- ER_per_person %>% mutate(marsh = if_else((UniqueIdentifier %in% MarshalleseUniqueID), 1,
                       if_else((UniqueIdentifier %in% ControlUniqueID), 0, NA) ) )
colnames(PCP_per_person)[3] <- "sum_PCP"
colnames(ER_per_person)[3] <- "sum_ER"
# 
length(unique(did_visit_types_year$UniqueIdentifier))
# # length is different than the nrow/2. Some not in both time periods? 
# 
# nrow(ER_per_person)
# nrow(PCP_per_person)
# # same
# length(unique(ER_per_person$UniqueIdentifier)) == length(unique(did_visit_types_year$UniqueIdentifier))
# 
# setdiff(ER_per_person$UniqueIdentifier, did_visit_types_year$UniqueIdentifier)
# # No difference
# 
# check <- did_visit_types_year %>% filter(UniqueIdentifier == 14012)


# did_visit_types_year_merge <- did_visit_types_year %>% select(UniqueIdentifier, marsh, pre_year)
# check <- left_join( did_visit_types_year_merge, ER_per_person,)
# ### Not working at all!!
```


```{r, include=FALSE}
gee_mod_DID_PCP_yr_best <- geeglm(sum_PCP ~ marsh*post_year, 
                             data = PCP_per_person,
                             id = UniqueIdentifier,
                             family = gaussian, 
                             scale.fix = T, # this sets phi = 1
                             corstr = "exchangeable")
#print(gee_mod_DID_PCP)
summary(gee_mod_DID_PCP_yr_best)

# Call:
#   geeglm(formula = sum_PCP ~ marsh * post_year, family = gaussian, 
#          data = PCP_per_person, id = UniqueIdentifier, corstr = "exchangeable", 
#          scale.fix = T)
# 
# Coefficients:
#   Estimate  Std.err    Wald Pr(>|W|)    
# (Intercept)      0.59361  0.00954 3872.74   <2e-16 ***
#   marsh           -0.12888  0.07324    3.10    0.078 .  
# post_year        0.14158  0.01293  119.86   <2e-16 ***
#   marsh:post_year -0.16577  0.08058    4.23    0.040 *  
#   ---
#   Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Correlation structure = exchangeable 
# Scale is fixed.
# 
# Link = identity 
# 
# Estimated Correlation Parameters:
#   Estimate Std.err
# alpha    0.208  0.0118
# Number of clusters:   19062  Maximum cluster size: 2
```

## PCP Regression Table

```{r, include=TRUE}
# gabby added titles here :)
tbl_regression(gee_mod_DID_PCP_yr_best, intercept = TRUE,
               label = list(
                 marsh ~ "Marshallese",
                 post_year ~ "Year")) %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  modify_table_styling(
    columns = label,
    rows = label == "Year",
    footnote = "Year relative to CHW hiring in 2019"
  )
```


```{r, include=FALSE}
gee_mod_DID_ER_yr_best <- geeglm(sum_ER ~ marsh*post_year, 
                                  data = ER_per_person,
                                  id = UniqueIdentifier,
                                  family = gaussian, 
                                  scale.fix = T, # this sets phi = 1
                                  corstr = "exchangeable")
#print(gee_mod_DID_PCP)
summary(gee_mod_DID_ER_yr_best)
coef(gee_mod_DID_ER_yr_best)
# confint(gee_mod_DID_ER_yr_best)

# geeglm(formula = sum_ER ~ marsh * post_year, family = gaussian, 
#        data = ER_per_person, id = UniqueIdentifier, corstr = "exchangeable", 
#        scale.fix = T)
# 
# Coefficients:
#   Estimate  Std.err    Wald Pr(>|W|)    
# (Intercept)      0.25948  0.00651 1589.51   <2e-16 ***
#   marsh            0.03425  0.04762    0.52    0.472    
# post_year        0.00164  0.00824    0.04    0.843    
# marsh:post_year -0.13055  0.05187    6.33    0.012 *  
#   ---
#   Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Correlation structure = exchangeable 
# Scale is fixed.
# 
# Link = identity 
# 
# Estimated Correlation Parameters:
#   Estimate Std.err
# alpha    0.282    0.03
# Number of clusters:   19062  Maximum cluster size: 2 


# GTSummary table of results


# reg2 <- glm(Affected ~ DHW.MeanMax_Degree_Heating_Weeks_MO03 + DHW.MeanMax_Degree_Heating_Weeks_YR10YR01 + mean_weekly_range_SST_CRW_Daily_ALLB4 + Prop_BleachResTaxa + Depth_m_mn,
#             family=poisson,
#             data = hcbc,
#             offset = log(LiveCoral))
# reg2
# 
# tbl_regression(reg2, exponentiate=TRUE, label=list(
#   DHW.MeanMax_Degree_Heating_Weeks_MO03 ~ 'Maximum DHW (prior 3 months)',
#   DHW.MeanMax_Degree_Heating_Weeks_YR10YR01 ~ 'Maximum DHW (prior 10 years)',
#   mean_weekly_range_SST_CRW_Daily_ALLB4 ~ 'Mean Weekly SST Range',
#   Prop_BleachResTaxa ~ 'Proportion of Bleach Resistant Taxa',
#   Depth_m_mn ~ 'Bathymetric Depth (m)'
# ))
# confint(gee_mod_DID_ER_yr_best)

# gabby added titles here :)
```

## ER Regression Table

```{r, include=TRUE}
tbl_regression(gee_mod_DID_ER_yr_best, intercept = TRUE,
               label = list(
                 marsh ~ "Marshallese",
                 post_year ~ "Year")) %>%
  bold_labels() %>%
  modify_header(label = "**Variable**") %>%
  modify_table_styling(
    columns = label,
    rows = label == "Year",
    footnote = "Year relative to CHW hiring in 2019")

```

# For ER Rates Population per Year

## All Patients in 2019 to 2022

Since the first Marshallese Community Health Worker (CHW) was hired in September 2019, we used the period of August 31, 2018 to August 31, 2019 as the pre-treatment populations of Marshallese and Non-Hispanic White patients at the Maple and Market clinics. We also used the same dates in 2021 and 2022 to find the post-treatment populations.

## Size of populations 
Marshallese `r length(unique(MarshalleseUniqueID))`

476

Non-Hispanic White `r length(unique(ControlUniqueID))`

18586

The population of CHAS patients has grown over time. Using data from all of the patients who visited the CHAS Maple and Market clinics during this time, we built a model that accounts for the correlation that results from patients who have been there in both 2019 and 2022, and includes patients who were not there in 2019 but joined by 2022. 

We modeled the number of ER visits between June 1 and August 31, 2019, and the number for the same dates in 2022, using a model with indicators for Marshallese, year, and the interaction of these 2 variables.

Using GEE GLM, we estimate that the expected value of ER visits per patients for our control population of  Non-Hispanic White patients in 2019 is `r round(gee_mod_DID_ER_yr_best$coefficients[1], 4 )`.

Similarly, we estimate the rate of ER visits per person visits for Marshallese in 2019 is `r round(gee_mod_DID_ER_yr_best$coefficients[1] + gee_mod_DID_ER_yr_best$coefficients[2], 4)`. This is slightly higher than the Non-Hispanic White population, but we note that the p-value not significant since it is above the alpha = 0.05 level (p-value = 0.5.)



Over the next three years, the rate of change for the ER visits Non-Hispanic White patients went up by `r round(gee_mod_DID_ER_yr_best$coefficients[3], 4)` per year. We note that this data was recorded during the COVID pandemic. This predicts no change in the average ER rates for the NHW group.

However, we note that the rate of change for the Marshallese patients goes down during the same time, `r round(gee_mod_DID_ER_yr_best$coefficients[3] + gee_mod_DID_ER_yr_best$coefficients[4], 4)` per year. The p-values for this coefficients is significant at 0.012. 

Based on the Difference in Difference causal model, we interpret lower rate of ER visits for the Marshallese as the average treatment effect of the Marshallese Community Health Workers. 


(We can use causal language with DID.)
! do we like our interpretations? Should I calculate the slope *3 for the 3 year change? 




# For Primary Care Provider (PCP) Visits Rates per Year
## All Patients in 2019 to 2022
Again, using a GEE GLM model, we estimate that the average Primary Care Provider visit rate for the Non-Hispanic White group in 2019 is `r gee_mod_DID_PCP_yr_best$ coefficients [1]`. 

The average rate for the Marshallese in 2019 is slightly lower at `r gee_mod_DID_PCP_yr_best$ coefficients [1] + gee_mod_DID_PCP_yr_best$ coefficients [2]` but with a p-value =0.078 it was not statistically different. 

Each year, the PCP rate went up slightly for the Non-Hispanic White group by `r gee_mod_DID_PCP_yr_best$ coefficients [3]`, but it went down very slightly for the Marshallese, `r gee_mod_DID_PCP_yr_best$ coefficients [3] + gee_mod_DID_PCP_yr_best$ coefficients [4]`. 
Both of these trends have significant p-values. 

Using similar assumptions for the ER section, this model estimates the average treatment effect of the Marshallese Community Health Workers as causing a slightly negative effect on Primary Care Visits. We give more context and interesting data for future research that may explain this unexpected result in our discussion section. 