---
title: "Capstone EDA"
author: "Gabby Lopez"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(gtsummary)
library(lubridate)
library(naniar)
library(flextable)
library(sjlabelled)
library(gt)
```

# Descriptive Statistics

```{r cars, include=FALSE, echo=FALSE}
a1c <- read.csv("Raw Data/UWDataA1cs.csv")
bmi <- read.csv("Raw Data/UWDataBMIs.csv")
bp <- read.csv("Raw Data/UWDataBP.csv")
diag <- read.csv("Raw Data/UWDataDiagnoses.csv")
enc <- read.csv("Raw Data/UWDataEncounters.csv")
panel <- read.csv("Raw Data/UWDataPanel.csv")
koh.attend <- read.csv("Raw Data/UWDataKOHAttendance.csv")
enc.extra <- read.csv("Raw Data/UWExtraEncounterData.csv")
```

Changing any cases where PreDM=1 and T2DM=1 to PreDM=0 for the descriptive statistics
```{r, include=FALSE, echo=FALSE}
diag$PreDM[diag$PreDM == 1 & diag$T2DM == 1] <- 0
```

```{r, include=FALSE, echo=FALSE}
# change date variable to date type and remove NA values in necessary fields
a1c$Date <- mdy(a1c$Date)
a1c.nona <- a1c %>% filter(!is.na(A1c) & !is.na(Date))

bmi$Date <- mdy(bmi$Date)
bmi.nona <- bmi %>% filter(!is.na(Date) & !is.na(BMI))

bp$Date <- mdy(bp$Date)
bp.nona <- bp %>% filter(!is.na(Date) & !is.na(Systolic))

enc <- bind_rows(enc, enc.extra)
enc$Date <- mdy(enc$Date)
enc.nona <- enc %>% filter(!is.na(Date) & !is.na(ServiceLine))

koh.attend$Date <- mdy(koh.attend$dateAttended)
koh.attend <- koh.attend %>% filter(!is.na(Date))
koh.counts <- koh.attend %>% group_by(UniqueIdentifier) %>% count(name = "koh.counts")

diag.nona <- diag %>% filter(!is.na(HTN) & !is.na(PreDM) & !is.na(T2DM))
# create diabetes indicator for both pre and T2DM and indicator for if they have both htn and diabetes
diag.nona <- diag.nona %>% mutate(Diabetes = ifelse((PreDM == 1 | T2DM == 1), 1, 0),
                                  both = ifelse((Diabetes == 1 & HTN == 1), 1, 0))
```

```{r, include=FALSE, echo=FALSE}
# Remove anyone under the age of 18
age <- panel %>% select(c("UniqueIdentifier","age"))
a1c.nona.18 <- left_join(a1c.nona, age, by = "UniqueIdentifier")
a1c.nona.18 <- a1c.nona.18 %>% filter(age >= 18)

bmi.nona.18 <- left_join(bmi.nona, age, by = "UniqueIdentifier")
bmi.nona.18 <- bmi.nona.18 %>% filter(age >= 18)
bmi.nona.18 <- bmi.nona.18 %>% select(-age)

bp.nona.18 <- left_join(bp.nona, age, by = "UniqueIdentifier")
bp.nona.18 <- bp.nona.18 %>% filter(age >= 18)

enc.nona.18 <- left_join(enc.nona, age, by = "UniqueIdentifier")
enc.nona.18 <- enc.nona.18 %>% filter(age >= 18)

diag.nona.18 <- left_join(diag.nona, age, by = "UniqueIdentifier")
diag.nona.18 <- diag.nona.18 %>% filter(age >= 18)

koh.attend.18 <- left_join(koh.attend, age, by = "UniqueIdentifier")
koh.attend.18 <- koh.attend.18 %>% filter(age >= 18)

# patients with KOHParticipant = 1 but are not in KOH attendance dataset
panel.koh <- panel %>% filter(KOHParticipant == 1 & age>=18)
k <- full_join(koh.counts, panel.koh, by = "UniqueIdentifier")
k %>% filter(is.na(koh.counts))

panel <- panel %>% mutate(Sex.long = case_when(Sex == 'M' ~ 'Male',
                                               Sex == 'F' ~ 'Female'))
panel.18 <- panel %>% filter(age >= 18)

# just checking the possible races
unique(panel.18$Race)
marsh <- panel.18 %>% filter(Race == 'Marshallese')
lang <- panel.18 %>% filter(Language == 'Marshallese')
#there are a lot of patients with Race = Marshallese that don't have language Marshallese and vice versa

# create our population of non-hispanic whites and marshallese (language = marsh or race = marsh or KOH = 1) - think there might be a KOH participant that doesn't have race or language as marshallese but I need to double check
targetpop <- panel.18 %>% filter(Race == 'Marshallese' | Language == 'Marshallese' | KOHParticipant == 1 | (Race == 'White' & Ethnicity == 'Not Hispanic or Latino'))

# remove any homeless population
targetpop <- targetpop %>% filter(No.column.name != 'Homeless Shelter' & No.column.name != 'Street')

# make marshallese indicator variable
targetpop <- targetpop %>% 
  mutate(Marsh = ifelse((Race == 'Marshallese' | Language == 'Marshallese' | KOHParticipant == 1), 1, 0),
         Group = ifelse(Marsh == 1, "Marshallese", "Non-Marshallese"))
```

```{r include=FALSE}
# Prep for table 1
unique(targetpop$No.column.name)
targetpop$No.column.name <- factor(targetpop$No.column.name, 
                                   levels = c('Housed','Doubling Up','Permanent Supportive Housing','Transitional','Other','Unknown'))

typeof(targetpop$Language)
unique(targetpop$ClinicLocation)

# replace all the blanks in the character variables with NAs so we can count the missings in the table
targetpop <- targetpop %>% replace_with_na_all(condition = ~.x == "")

# get rid of the age variable we got from merging to filter out children since targetpop already has age
diag.nona.18 <- diag.nona.18 %>% select(-age)

# merge diagnosis data with other characteristics
table1 <- right_join(diag.nona.18, targetpop, by = "UniqueIdentifier")

# create subset of only KOH participants
koh <- table1 %>% filter(KOHParticipant == 1)
koh.table1 <- inner_join(koh.counts, koh, by = "UniqueIdentifier")
```

distribution of KOH attendance
```{r echo=FALSE}
hist(koh.table1$koh.counts)
summary(koh.table1$koh.counts)
summary(as.factor(koh.table1$koh.counts))
```
## Primary Objective - Hypertension pre and post analysis datasets ##
subsetting koh patients with hypertension
```{r include=FALSE}
# get pts with htn
koh.htn <- koh.table1 %>% filter(HTN == 1)
# isolate the patient ids
koh.htn.patlist <- koh.htn %>% select("UniqueIdentifier")
# select bp measures for only kohn htn pts
koh.bp <- left_join(koh.htn.patlist, bp.nona.18, by = "UniqueIdentifier") %>% select(-age)

# count number of bp readings per patient
koh.bp.counts <- koh.bp %>% group_by(UniqueIdentifier) %>% count(name = "bp.counts")
# remove any patients that only have 1 reading - i will filter out any that don't have reading before and after their KOH meeting but for now, we at least know they need to have at least 2 bp readings
koh.htn.elig.patlist <- koh.bp.counts %>% filter(bp.counts > 1) %>% select("UniqueIdentifier")
# now that we have the "eligible patients" we can subset the BP data for these patients
koh.bp.elig <- left_join(koh.htn.elig.patlist, koh.bp, by = "UniqueIdentifier")


# getting the date of first visit attended
koh.mtg1 <- koh.attend.18 %>% group_by(UniqueIdentifier) %>% arrange(Date, .by_group=TRUE) %>% slice_head() %>% select(c("UniqueIdentifier", "Date")) %>% rename(KOHDate = Date)

# merging with eligble bp table to find "pre" BP reading closest to first KOH meeting
koh.mtg1.bp <- right_join(koh.mtg1, koh.bp.elig, by="UniqueIdentifier") %>% 
  rename(BPDate = Date) %>%
  mutate(datediff = as.numeric(difftime(BPDate, KOHDate, units="days")))

# these are all the "pre" KOH BP measurements
koh.bp.pre <- koh.mtg1.bp %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, BPDate.pre = BPDate, sys.pre = Systolic, dia.pre = Diastolic)

avg.bmi <- bmi.nona.18 %>% group_by(UniqueIdentifier) %>% mutate(avg.bmi = mean(BMI)) %>% slice_head() %>% select(c(UniqueIdentifier, avg.bmi))

# now finding the most recent BP measure for "post" KOH
koh.bp.post <- koh.mtg1.bp %>% filter(datediff>0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, BPDate.post = BPDate, sys.post = Systolic, dia.post = Diastolic)

# now merging together the pre and post datasets to find patients that have both timepoints
koh.bp.pre.post <- inner_join(koh.bp.pre, koh.bp.post, by="UniqueIdentifier")
```

subsetting marshallese non-koh pts with HTN
```{r include=FALSE}
marsh.non.htn <- table1 %>% filter(Marsh == 1 & KOHParticipant == 0 & HTN == 1)
# isolate the patient ids
marsh.non.htn.patlist <- marsh.non.htn %>% select("UniqueIdentifier")
# select bp measures for only non koh htn pts
marsh.non.bp <- left_join(marsh.non.htn.patlist, bp.nona.18, by = "UniqueIdentifier") %>% select(-age)

# count number of bp readings per patient
marsh.non.bp.counts <- marsh.non.bp %>% group_by(UniqueIdentifier) %>% count(name = "bp.counts")
# remove any patients that only have 1 reading - i will filter out any that don't have reading before and after first KOH meeting but for now, we at least know they need to have at least 2 bp readings
marsh.non.htn.elig.patlist <- marsh.non.bp.counts %>% filter(bp.counts > 1) %>% select("UniqueIdentifier")
# now that we have the "eligible patients" we can subset the BP data for these patients
marsh.non.bp.elig <- left_join(marsh.non.htn.elig.patlist, marsh.non.bp, by = "UniqueIdentifier")

# the earliest KOH meeting in the koh subsets is 4/5/2023, so we will use that as our starting point to find "pre" readings
marsh.non.bp.elig$KOHDate <- as.Date('2023-04-05')
# find "pre" BP reading closest to first KOH meeting
marsh.non.bp.elig <- marsh.non.bp.elig %>% rename(BPDate = Date) %>%
  mutate(datediff = as.numeric(difftime(BPDate, KOHDate, units="days")))

# these are all the "pre" KOH BP measurements
marsh.non.bp.pre <- marsh.non.bp.elig %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, BPDate.pre = BPDate, sys.pre = Systolic, dia.pre = Diastolic)

# now finding the most recent BP measure for "post" KOH
marsh.non.bp.post <- marsh.non.bp.elig %>% filter(datediff>0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, BPDate.post = BPDate, sys.post = Systolic, dia.post = Diastolic)

# now merging together the pre and post datasets to find patients that have both timepoints
marsh.non.bp.pre.post <- inner_join(marsh.non.bp.pre, marsh.non.bp.post, by="UniqueIdentifier")

# just looking at the pre and post means between groups
mean(marsh.non.bp.pre.post$sys.pre)
mean(marsh.non.bp.pre.post$sys.post)
mean(koh.bp.pre.post$sys.pre)
mean(koh.bp.pre.post$sys.post)

marsh.non.bp.pre.post <- marsh.non.bp.pre.post %>% select(-c(KOH.start.dt.x, KOH.start.dt.y))
marsh.non.bp.pre.post_long <- marsh.non.bp.pre.post %>%
  pivot_longer(
    cols = c(BPDate.pre, sys.pre, dia.pre, BPDate.post, sys.post, dia.post),
    names_to = c(".value", "TimePoint"),
    names_sep = "\\."
  ) %>%
  select(UniqueIdentifier, BPDate, sys, dia)

koh.bp.pre.post <- koh.bp.pre.post %>% select(-c(KOH.start.dt.x, KOH.start.dt.y))
koh.bp.pre.post_long <- koh.bp.pre.post %>%
  pivot_longer(
    cols = c(BPDate.pre, sys.pre, dia.pre, BPDate.post, sys.post, dia.post),
    names_to = c(".value", "TimePoint"),
    names_sep = "\\."
  ) %>%
  select(UniqueIdentifier, BPDate, sys, dia)
koh.bp.pre.post_long$KOH <- 1
obj1.bp.pre.post <- bind_rows(koh.bp.pre.post_long, marsh.non.bp.pre.post_long)
obj1.bp.pre.post$KOH[is.na(obj1.bp.pre.post$KOH)] <- 0
obj1.bp.pre.post <- left_join(obj1.bp.pre.post, koh.counts, by = "UniqueIdentifier")
obj1.bp.pre.post$koh.counts[is.na(obj1.bp.pre.post$koh.counts)] <- 0
# discretizing koh attendance
obj1.bp.pre.post <- obj1.bp.pre.post %>% mutate(KOH.none = ifelse(koh.counts == 0, 1, 0),
                                                KOH.one = ifelse(koh.counts == 1, 1, 0),
                                                KOH.mult = ifelse(koh.counts > 1, 1, 0))
obj1.bp.cov <- table1 %>% select(c(UniqueIdentifier, age, Sex, IncomeLevel, BLACERISK))
obj1.bp.pre.post.full <- left_join(obj1.bp.pre.post, obj1.bp.cov, by="UniqueIdentifier")
obj1.bp.pre.post.full <- left_join(obj1.bp.pre.post.full, avg.bmi, by="UniqueIdentifier")

write.csv(obj1.bp.pre.post.full, 'Obj1BPPrePost.csv')
```

## Primary objective hypertension longitudinal analysis datasets ##
koh patients
```{r include=FALSE}
# these are all the "pre" KOH BP measurements
koh.bp.pre.all <- koh.mtg1.bp %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)

# now finding all BP measures for "post" KOH
koh.bp.post.all <- koh.mtg1.bp %>% filter(datediff>0) %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)
# combining baseline and subsequent measurements
koh.bp.all <- bind_rows(koh.bp.pre.all, koh.bp.post.all)
# making sure final list has both a pre and post timepoint
koh.bp.pre.post.pt <- koh.bp.pre.post %>% select(UniqueIdentifier)
koh.bp.all.elig <- left_join(koh.bp.pre.post.pt, koh.bp.all, by = "UniqueIdentifier")
koh.bp.all.elig$KOH <- 1
```

marshallese non-koh patients
```{r include=FALSE}
# these are all the "pre" KOH BP measurements
marsh.non.bp.pre.all <- marsh.non.bp.elig %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)

# now finding all BP measures for "post" KOH
marsh.non.bp.post.all <- marsh.non.bp.elig %>% filter(datediff>0) %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)
# combining baseline and subsequent measurements
marsh.non.bp.all <- bind_rows(marsh.non.bp.pre.all, marsh.non.bp.post.all)
# making sure final list has both a pre and post timepoint
marsh.non.bp.pre.post.pt <- marsh.non.bp.pre.post %>% select(UniqueIdentifier)
marsh.non.bp.all.elig <- left_join(marsh.non.bp.pre.post.pt, marsh.non.bp.all, by = "UniqueIdentifier")
marsh.non.bp.all.elig$KOH <- 0

# combining koh with non-koh
obj1.bp.lme <- bind_rows(koh.bp.all.elig, marsh.non.bp.all.elig)
obj1.bp.lme <- left_join(obj1.bp.lme, koh.counts, by = "UniqueIdentifier")
obj1.bp.lme$koh.counts[is.na(obj1.bp.lme$koh.counts)] <- 0
# discretizing koh attendance
obj1.bp.lme <- obj1.bp.lme %>% mutate(KOH.none = ifelse(koh.counts == 0, 1, 0),
                                                KOH.one = ifelse(koh.counts == 1, 1, 0),
                                                KOH.mult = ifelse(koh.counts > 1, 1, 0))
obj1.bp.lme.full <- left_join(obj1.bp.lme, obj1.bp.cov, by="UniqueIdentifier")
obj1.bp.lme.full <- left_join(obj1.bp.lme.full, avg.bmi, by="UniqueIdentifier")

write.csv(obj1.bp.lme.full, 'Obj1BP_LME.csv')
```

## Primary Objective - Diabetes pre and post analysis datasets ##
subsetting koh patients with prediabetes or T2DM
```{r include=FALSE}
# get pts with diabetes
koh.t2dm <- koh.table1 %>% filter(Diabetes == 1)
# isolate the patient ids
koh.t2dm.patlist <- koh.t2dm %>% select("UniqueIdentifier")
# select a1c measures for only koh t2dm pts
koh.a1c <- left_join(koh.t2dm.patlist, a1c.nona.18, by = "UniqueIdentifier") %>% select(-age)

# count number of a1c readings per patient
koh.a1c.counts <- koh.a1c %>% group_by(UniqueIdentifier) %>% count(name = "a1c.counts")
# remove any patients that only have 1 reading - i will filter out any that don't have reading before and after their KOH meeting but for now, we at least know they need to have at least 2 a1c readings
koh.t2dm.elig.patlist <- koh.a1c.counts %>% filter(a1c.counts > 1) %>% select("UniqueIdentifier")
# now that we have the "eligible patients" we can subset the a1c data for these patients
koh.a1c.elig <- left_join(koh.t2dm.elig.patlist, koh.a1c, by = "UniqueIdentifier")


# find "pre" A1c reading closest to first KOH meeting
koh.mtg1.a1c <- right_join(koh.mtg1, koh.a1c.elig, by="UniqueIdentifier") %>% 
  rename(A1cDate = Date) %>%
  mutate(datediff = as.numeric(difftime(A1cDate, KOHDate, units="days")))

# these are all the "pre" KOH A1c measurements
koh.a1c.pre <- koh.mtg1.a1c %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, A1cDate.pre = A1cDate, A1c.pre = A1c)

# now finding the most recent A1c measure for "post" KOH
koh.a1c.post <- koh.mtg1.a1c %>% filter(datediff>0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, A1cDate.post = A1cDate, A1c.post = A1c)

# now merging together the pre and post datasets to find patients that have both timepoints
koh.a1c.pre.post <- inner_join(koh.a1c.pre, koh.a1c.post, by="UniqueIdentifier")
```

subsetting marshallese non-koh pts with diabetes
```{r include=FALSE}
marsh.non.dia <- table1 %>% filter(Marsh == 1 & KOHParticipant == 0 & Diabetes == 1)
# isolate the patient ids
marsh.non.dia.patlist <- marsh.non.dia %>% select("UniqueIdentifier")
# select bp measures for only non koh diabetic pts
marsh.non.a1c <- left_join(marsh.non.dia.patlist, a1c.nona.18, by = "UniqueIdentifier") %>% select(-age)

# count number of a1c readings per patient
marsh.non.a1c.counts <- marsh.non.a1c %>% group_by(UniqueIdentifier) %>% count(name = "a1c.counts")
# remove any patients that only have 1 reading - i will filter out any that don't have reading before and after first KOH meeting but for now, we at least know they need to have at least 2 a1c readings
marsh.non.dia.elig.patlist <- marsh.non.a1c.counts %>% filter(a1c.counts > 1) %>% select("UniqueIdentifier")
# now that we have the "eligible patients" we can subset the A1c data for these patients
marsh.non.a1c.elig <- left_join(marsh.non.dia.elig.patlist, marsh.non.a1c, by = "UniqueIdentifier")

# the earliest KOH meeting in the koh subsets is 4/5/2023, so we will use that as our starting point to find "pre" readings
marsh.non.a1c.elig$KOHDate <- as.Date('2023-04-05')
# find "pre" A1c reading closest to first KOH meeting
marsh.non.a1c.elig <- marsh.non.a1c.elig %>% rename(A1cDate = Date) %>%
  mutate(datediff = as.numeric(difftime(A1cDate, KOHDate, units="days")))

# these are all the "pre" KOH A1c measurements
marsh.non.a1c.pre <- marsh.non.a1c.elig %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, A1cDate.pre = A1cDate, A1c.pre = A1c)

# now finding the most recent A1c measure for "post" KOH
marsh.non.a1c.post <- marsh.non.a1c.elig %>% filter(datediff>0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate, A1cDate.post = A1cDate, A1c.post = A1c)

# now merging together the pre and post datasets to find patients that have both timepoints
marsh.non.a1c.pre.post <- inner_join(marsh.non.a1c.pre, marsh.non.a1c.post, by="UniqueIdentifier")

# just looking at the pre and post means between groups
mean(marsh.non.a1c.pre.post$A1c.pre)
mean(marsh.non.a1c.pre.post$A1c.post)
mean(koh.a1c.pre.post$A1c.pre)
mean(koh.a1c.pre.post$A1c.post)

# PRIMARY OBJECTIVE A1C PRE-POST DATASET
marsh.non.a1c.pre.post <- marsh.non.a1c.pre.post %>% select(-c(KOH.start.dt.x, KOH.start.dt.y))
marsh.non.a1c.pre.post_long <- marsh.non.a1c.pre.post %>%
  pivot_longer(
    cols = c(A1cDate.pre, A1c.pre, A1cDate.post, A1c.post),
    names_to = c(".value", "TimePoint"),
    names_sep = "\\."
  ) %>%
  select(UniqueIdentifier, A1cDate, A1c)

koh.a1c.pre.post <- koh.a1c.pre.post %>% select(-c(KOH.start.dt.x, KOH.start.dt.y))
koh.a1c.pre.post_long <- koh.a1c.pre.post %>%
  pivot_longer(
    cols = c(A1cDate.pre, A1c.pre, A1cDate.post, A1c.post),
    names_to = c(".value", "TimePoint"),
    names_sep = "\\."
  ) %>%
  select(UniqueIdentifier, A1cDate, A1c)
koh.a1c.pre.post_long$KOH <- 1
obj1.a1c.pre.post <- bind_rows(koh.a1c.pre.post_long, marsh.non.a1c.pre.post_long)
obj1.a1c.pre.post$KOH[is.na(obj1.a1c.pre.post$KOH)] <- 0
obj1.a1c.pre.post <- left_join(obj1.a1c.pre.post, koh.counts, by = "UniqueIdentifier")
obj1.a1c.pre.post$koh.counts[is.na(obj1.a1c.pre.post$koh.counts)] <- 0
# discretizing koh attendance
obj1.a1c.pre.post <- obj1.a1c.pre.post %>% mutate(KOH.none = ifelse(koh.counts == 0, 1, 0),
                                                KOH.one = ifelse(koh.counts == 1, 1, 0),
                                                KOH.mult = ifelse(koh.counts > 1, 1, 0))
obj1.a1c.pre.post.full <- left_join(obj1.a1c.pre.post, obj1.bp.cov, by="UniqueIdentifier")
obj1.a1c.pre.post.full <- left_join(obj1.a1c.pre.post.full, avg.bmi, by="UniqueIdentifier")

write.csv(obj1.a1c.pre.post.full, 'Obj1A1cPrePost.csv')
```
## Primary objective Diabetes longitudinal analysis datasets ##
koh pts
```{r include=FALSE}
# these are all the "pre" KOH A1c measurements
koh.a1c.pre.all <- koh.mtg1.a1c %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)

# now finding all A1c measures for "post" KOH
koh.a1c.post.all <- koh.mtg1.a1c %>% filter(datediff>0) %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)
# combining baseline and subsequent measurements
koh.a1c.all <- bind_rows(koh.a1c.pre.all, koh.a1c.post.all)
# making sure final list has both a pre and post timepoint
koh.a1c.pre.post.pt <- koh.a1c.pre.post %>% select(UniqueIdentifier)
koh.a1c.all.elig <- left_join(koh.a1c.pre.post.pt, koh.a1c.all, by = "UniqueIdentifier")
koh.a1c.all.elig$KOH <- 1
```

marshallese non-koh pts
```{r include=FALSE}
# these are all the "pre" KOH A1c measurements
marsh.non.a1c.pre.all <- marsh.non.a1c.elig %>% filter(datediff<=0) %>% group_by(UniqueIdentifier) %>% arrange(desc(datediff), .by_group=TRUE) %>% slice_head() %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)

# now finding all A1c measures for "post" KOH
marsh.non.a1c.post.all <- marsh.non.a1c.elig %>% filter(datediff>0) %>% select(-(datediff)) %>%
  rename(KOH.start.dt = KOHDate)
# combining baseline and subsequent measurements
marsh.non.a1c.all <- bind_rows(marsh.non.a1c.pre.all, marsh.non.a1c.post.all)
# making sure final list has both a pre and post timepoint
marsh.non.a1c.pre.post.pt <- marsh.non.a1c.pre.post %>% select(UniqueIdentifier)
marsh.non.a1c.all.elig <- left_join(marsh.non.a1c.pre.post.pt, marsh.non.a1c.all, by = "UniqueIdentifier")
marsh.non.a1c.all.elig$KOH <- 0

obj1.a1c.lme <- bind_rows(koh.a1c.all.elig, marsh.non.a1c.all.elig)
obj1.a1c.lme <- left_join(obj1.a1c.lme, koh.counts, by = "UniqueIdentifier")
obj1.a1c.lme$koh.counts[is.na(obj1.a1c.lme$koh.counts)] <- 0
# discretizing koh attendance
obj1.a1c.lme <- obj1.a1c.lme %>% mutate(KOH.none = ifelse(koh.counts == 0, 1, 0),
                                                KOH.one = ifelse(koh.counts == 1, 1, 0),
                                                KOH.mult = ifelse(koh.counts > 1, 1, 0))
obj1.a1c.lme.full <- left_join(obj1.a1c.lme, obj1.bp.cov, by="UniqueIdentifier")
obj1.a1c.lme.full <- left_join(obj1.a1c.lme.full, avg.bmi, by="UniqueIdentifier")

write.csv(obj1.a1c.lme.full, 'Obj1A1c_LME.csv')
```

make table 1 dataset for primary objective 249 unique patients 127 from bp and 221 from a1c
```{r include=FALSE}
# combine eligible bp and a1c patients and get table 1 covariates
obj1 <- full_join(obj1.bp.pre.post, obj1.a1c.pre.post, 
                  by = c("UniqueIdentifier", "KOH","koh.counts","KOH.none","KOH.one","KOH.mult")) %>%
  select(c(UniqueIdentifier, KOH, koh.counts)) %>% slice_head()

obj1 <- left_join(obj1, table1, by = "UniqueIdentifier")
obj1 <- obj1 %>% mutate(KOH.cat = case_when(KOH == 1 ~ 'KOH Participant',
                                            KOH == 0 ~ 'Non-Participant'))
obj1 <- left_join(obj1, avg.bmi, by = "UniqueIdentifier")

```

```{r}
obj1 <- read.csv('Analysis Data/Obj1_AllPts.csv')[,-1]
```

table 1 for primary objective population
\pagebreak
```{r echo=FALSE, warning=FALSE}
obj1 %>% select(c('KOH.cat','age','Sex.long','avg.bmi','HTN','PreDM','T2DM','both','ClinicLocation','No.column.name','IncomeLevel','BLACERISK')) %>%
   gtsummary::tbl_summary(
     by = KOH.cat,
     include = c(age, Sex.long, avg.bmi, HTN, PreDM, T2DM, both, ClinicLocation, No.column.name, IncomeLevel, BLACERISK),
     label = list(
       age ~ 'Age',
       Sex.long ~ 'Sex',
       avg.bmi ~ 'BMI',
       HTN ~ 'Hypertension',
       PreDM ~ 'Pre-Diabetes',
       T2DM ~ 'Type II Diabetes',
       both ~ 'Hypertension and T2DM',
       ClinicLocation ~ 'Clinic Location',
       No.column.name ~ 'Housing Status',
       IncomeLevel ~ 'Income Level',
       BLACERISK ~ 'Risk Score'
     ),
     missing_text = "Missing",
     type = all_continuous() ~ "continuous",
     statistic = all_continuous() ~ c(
       "{median} ({min}, {max})"
     ),
     digits = all_continuous() ~ 1
   )  %>%
  bold_labels() %>%
  modify_table_styling(
    columns = label,
    rows = label == "Income Level",
    footnote = "Percentage of participant's annual income relative to the federal poverty line"
  ) %>%
  modify_table_styling(
    columns = label,
    rows = label == "Risk Score",
    footnote = "Risk scores are calculated based on the BLACE Index Scoring Tool. Scores range from 0-19; <8 is low risk, 8-11 is moderate risk, and 12+ is high risk."
  ) %>%
   modify_header(label ~ "**Baseline Characteristics**") %>%
  modify_header(all_stat_cols() ~ "**{level}**<br>\nN = {n}") %>%
  modify_caption("Marshallese Participant Characteristics") %>%
   as_flex_table()
```


table 1 for all Marshallese and non-Marshallese
```{r echo=FALSE, warning=FALSE}
obj2 <- read.csv('Analysis Data/Obj2_AllPts.csv')[,-1]
obj2 %>% select(c('Group','age','Sex.long','avg.bmi','HTN','PreDM','T2DM','both','ClinicLocation','No.column.name','IncomeLevel','BLACERISK')) %>%
   gtsummary::tbl_summary(
     by = Group,
     include=c(age, Sex.long, avg.bmi, HTN, PreDM, T2DM, both, ClinicLocation, No.column.name, IncomeLevel, BLACERISK),
     list(
       age ~ 'Age',
       Sex.long ~ 'Sex',
       avg.bmi ~ 'BMI',
       HTN ~ 'Hypertension',
       PreDM ~ 'Pre-Diabetes',
       T2DM ~ 'Type II Diabetes',
       both ~ 'Hypertension and T2DM',
       ClinicLocation ~ 'Clinic Location',
       No.column.name ~ 'Housing Status',
       IncomeLevel ~ 'Income Level',
       BLACERISK ~ 'Risk Score'
     ),
     missing_text = "Missing",
     type = all_continuous() ~ "continuous",
     statistic = all_continuous() ~ c(
       "{median} ({min}, {max})"
     ),
     digits = all_continuous() ~ 1
   ) %>%
   bold_labels() %>%
  modify_table_styling(
    columns = label,
    rows = label == "Income Level",
    footnote = "Percentage of participant's annual income relative to the federal poverty line."
  ) %>%
  modify_table_styling(
    columns = label,
    rows = label == "Risk Score",
    footnote = "Risk scores are calculated based on the BLACE Index Scoring Tool. Scores range from 0-19; <8 is low risk, 8-11 is moderate risk, and 12+ is high risk."
  ) %>%
   modify_header(label ~ "**Baseline Characteristics**") %>%
   modify_header(all_stat_cols() ~ "**{level}**<br>\nN = {n}") %>%
  modify_caption("Marshallese and Non-Hispanic White Participant Characteristics") %>%
   add_overall() %>%
   as_flex_table()
```

# One-Time KOH Attendance Distribution

```{r echo=FALSE}
koh_1x = koh.counts %>% filter(koh.counts == 1)
koh_1x = koh_1x %>% 
  left_join(koh.attend.18, by="UniqueIdentifier") %>% 
  select(Date)
koh_1x = koh_1x %>% 
  group_by(Date) %>% 
  summarize(attendees=n(), .groups="drop")

ggplot(koh_1x, aes(x=Date, y=attendees)) +
  geom_col(fill="steelblue") +
  geom_text(aes(label=Date, angle=90, vjust=-0.5)) +
  labs(title="Number of One-Time KOH Attendees by Date",
       x="KOH Meeting Date",
       y="Number of Attendees")
```


# Preliminary Parallel Trends EDA

```{r echo=FALSE}
# join encounter and panel data to get visit types by race
enc_wraces = enc.nona.18 %>% 
  left_join(panel.18, by="UniqueIdentifier") %>% 
  select(Date, ServiceLine, Race, Language)

# dates up to March 2023 are of interest, also t_0 to new yr's day 2021
tot_months = 26
t_0 = enc_wraces$Date %>% min() - days(8)

# initialize empty dataframe to store monthly PC visit counts
pc_trends = tibble(
  visits = numeric(),
  race = character(),
  Date = Date()
)

# iterate through all months and count PC visits
curr_date = t_0
for (i in 1:tot_months) {
  # filter for only PC visits
  curr_pc_visits = enc_wraces %>% filter(
    month(Date) == month(curr_date) &
    year(Date) == year(curr_date) & 
    ServiceLine == "Primary Care"
  )
  
  # filter for/add visits by Marshallese
  curr_marsh_visits = curr_pc_visits %>% filter(
    Race == "Marshallese" |
    Language == "Marshallese"
  ) %>% count()
  
  pc_trends = pc_trends %>% add_row(
    # increase Marshallese visits counts by 2 orders of magnitude for easy comparison
    # TODO: REVERT REVERT REVERT
    visits = curr_marsh_visits[1, 1] * 100,
    race = "Marshallese",
    Date = make_date(year=year(curr_date), month=month(curr_date), day=1)
  )
  
  # filter for/add visits by NHWs
  curr_nhwht_visits = curr_pc_visits %>% filter(
    Race == "White"
  ) %>% count()
  
  pc_trends = pc_trends %>% add_row(
    visits = curr_nhwht_visits[1, 1],
    race = "Non-Hispanic White",
    Date = make_date(year=year(curr_date), month=month(curr_date), day=1)
  )
  
  # increment date by 1 month
  curr_date = curr_date + months(1)
}

# plot
ggplot(pc_trends, aes(x=Date, y=visits, color=race, group=race)) +
  geom_line() +
  geom_point() + 
  labs(
    title="Primary Care Visits by Race",
    x="Date",
    y="Number of Visits",
    color="Race"
  )
```

## Encounters Table ##
```{r include=FALSE}
# getting the dataset in the proper form

pop.race <- targetpop %>% select(c('UniqueIdentifier','Group'))
# add race to encounters table
enc.race <- inner_join(pop.race, enc.nona.18, by='UniqueIdentifier')

# isolate year to use in table
enc.race$year <- as.numeric(format(enc.race$Date, "%Y"))
enc.race$ServiceLine[enc.race$ServiceLine == "Unmapped"] <- "Other"
enc.race$ServiceLine <- factor(enc.race$ServiceLine,
                               levels = c("Primary Care", "Emergency", "Urgent Care", "Behavioral Health",
                                          "Pharmacy", "Health Equity", "Dental", "Multispecialty", "Other"))
# calculate number of type of visit per year for each patient
enc.counts <- enc.race %>% group_by(UniqueIdentifier, year, ServiceLine) %>% count(name = "enc.counts")
enc.counts.race <- inner_join(pop.race, enc.counts, by='UniqueIdentifier')
enc.counts.race$Group <- factor(enc.counts.race$Group, levels = c("Marshallese","Non-Marshallese"))
enc.counts.wide <- enc.counts.race %>%
  pivot_wider(
    names_from = ServiceLine,
    values_from = enc.counts,
    values_fill = 0 # Fill missing values with 0
  )
enc.counts.wide <- enc.counts.wide %>% rename(PrimaryCare = 'Primary Care', UrgentCare = 'Urgent Care',
                                              Behavioral = 'Behavioral Health', Equity = 'Health Equity') %>%
  mutate(across(c(PrimaryCare, Emergency, UrgentCare, Behavioral, Pharmacy, 
                  Equity, Dental, Multispecialty, Other), as.numeric))
```
i think i still need to adjust this table a bit but it's fine for now. I removed visit types (Health Equity and Other) that had 0 visits for multiple years as well. 2016, 2017, and 2025 didn't have much data either so I removed them. Then I removed years 2018 and 2019 just to make the table a bit smaller since it was so wide.
```{r echo=FALSE}
# making the encounters table
str(enc.counts.wide)
enc.marsh <- enc.counts.wide %>% filter(Group=='Marshallese' & year != "2016" & year != "2017" & year != "2018" & year != "2019" & year != "2025") %>%
  select(year, PrimaryCare, Emergency, UrgentCare, Behavioral, Pharmacy, Dental, Multispecialty) %>%
  mutate(across(c(PrimaryCare, Emergency, UrgentCare, Behavioral, Pharmacy, 
                  Dental, Multispecialty), as.numeric)) %>%
        gtsummary::tbl_summary(
         by = year,
        include = c(PrimaryCare, Emergency, UrgentCare, Behavioral, Pharmacy, Dental, Multispecialty),
        label = list(
          PrimaryCare ~ 'Primary Care',
          Emergency ~ 'Emergency',
          UrgentCare ~ 'Urgent Care',
          Behavioral ~ 'Behavioral Health',
          Pharmacy ~ 'Pharmacy',
          Dental ~ 'Dental',
          Multispecialty ~ 'Multispecialty'
        ),
        missing="no",
        type = list(PrimaryCare ~ "continuous", Emergency ~ "continuous", UrgentCare ~ "continuous", Behavioral ~ "continuous", Pharmacy ~ "continuous", Dental ~ "continuous", Multispecialty ~ "continuous"),
        statistic = all_continuous() ~ c("{mean} ({min}, {max})"),
        digits = all_continuous() ~ 1
      ) %>%
  bold_labels() %>%
  modify_footnote(c(all_stat_cols()) ~ "Mean (Minimum, Maximum)") %>%
  modify_header(label ~ "**Mean Visits per Patient**") %>%
  modify_header(all_stat_cols() ~ "**{level}**<br>\nN = {n}")

enc.nhw <- enc.counts.wide %>% filter(Group=='Non-Marshallese' & year != "2016" & year != "2017" & year != "2018" & year != "2019" & year != "2025") %>%
  select(year, PrimaryCare, Emergency, UrgentCare, Behavioral, Pharmacy, Dental, Multispecialty) %>%
  mutate(across(c(PrimaryCare, Emergency, UrgentCare, Behavioral, Pharmacy, 
                  Dental, Multispecialty), as.numeric)) %>%
        gtsummary::tbl_summary(
         by = year,
        include = c(PrimaryCare, Emergency, UrgentCare, Behavioral, Pharmacy, Dental, Multispecialty),
        label = list(
          PrimaryCare ~ 'Primary Care',
          Emergency ~ 'Emergency',
          UrgentCare ~ 'Urgent Care',
          Behavioral ~ 'Behavioral Health',
          Pharmacy ~ 'Pharmacy',
          Dental ~ 'Dental',
          Multispecialty ~ 'Multispecialty'
        ),
        missing="no",
        type = list(PrimaryCare ~ "continuous", Emergency ~ "continuous", UrgentCare ~ "continuous", Behavioral ~ "continuous", Pharmacy ~ "continuous", Dental ~ "continuous", Multispecialty ~ "continuous"),
        statistic = all_continuous() ~ c("{mean} ({min}, {max})"),
        digits = all_continuous() ~ 1
      ) %>%
  bold_labels() %>%
  modify_footnote(c(all_stat_cols()) ~ "Mean (Minimum, Maximum)") %>%
  modify_header(label ~ "**Mean Visits per Patient**") %>%
  modify_header(all_stat_cols() ~ "**{level}**<br>\nN = {n}")

enc.tbl <- tbl_merge(tbls = list(enc.marsh, enc.nhw), tab_spanner = c("**Marshallese**", "**Non-Marshallese**")) %>% modify_caption("**Clinic Visits Over Time**")
enc.tbl
```

```{r}
obj1.koh <- obj1 %>% filter(KOH==1)
obj1.koh$koh.counts.factor <- factor(obj1.koh$koh.counts)
obj1.koh$koh.counts <- as.numeric(obj1.koh$koh.counts)

obj1.koh.bp <- obj1.koh %>% filter(HTN==1)
koh.bp.tbl <- obj1.koh.bp %>% 
select(c("koh.counts","koh.counts.factor")) %>%
   gtsummary::tbl_summary(
     include=c(koh.counts, koh.counts.factor),
     label = list(
       koh.counts ~ 'Average KOH Meetings Attended',
       koh.counts.factor ~ 'KOH Meetings Attended'
     ),
     missing_text = "Missing",
     type = list(koh.counts ~ "continuous"),
     statistic = all_continuous() ~ c(
       "{median} ({min}, {max})"
     ),
     digits = all_continuous() ~ 1
   ) %>%
   bold_labels() %>%
   modify_header(all_stat_cols() ~ "**Hypertension**<br>\nN = 42")

obj1.koh.a1c <- obj1.koh %>% filter(Diabetes==1)
koh.a1c.tbl <- obj1.koh.a1c %>% 
select(c("koh.counts","koh.counts.factor")) %>%
   gtsummary::tbl_summary(
     include=c(koh.counts, koh.counts.factor),
     label = list(
       koh.counts ~ 'Average KOH Meetings Attended',
       koh.counts.factor ~ 'KOH Meetings Attended'
     ),
     missing_text = "Missing",
     type = list(koh.counts ~ "continuous"),
     statistic = all_continuous() ~ c(
       "{median} ({min}, {max})"
     ),
     digits = all_continuous() ~ 1
   ) %>%
   bold_labels() %>%
   modify_header(all_stat_cols() ~ "**Diabetes**<br>\nN = 71")

koh.tbl <- tbl_merge(tbls = list(koh.bp.tbl, koh.a1c.tbl), tab_spanner = FALSE) %>% modify_caption("**'Know Our Health' Participation by Disease Type**")
koh.tbl 
```

