---
title: "Objective 1 Longitudinal Analysis"
author: "Gabby Lopez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)
library(gtsummary)
library(gt)
library(kableExtra)
```
note to self: run regression diagnostics and ignore clustering - regular linear regression and see if any observations are having huge impact
change days to months

83 none; 14 one; 23 multiple; 120 total
```{r}
bp <- read.csv('Analysis Data/Obj1BP_LME.csv')[,-1]
```

```{r}
# standardizing time as days from baseline measurement
bp.first <- bp %>% group_by(UniqueIdentifier) %>% arrange(BPDate, .by_group=TRUE) %>% slice_head()
bp.first$BPDate <- bp.first$KOH.start.dt
bp <- bp %>% filter(BPDate>KOH.start.dt)

bp <- rbind(bp.first, bp) %>% arrange(UniqueIdentifier, BPDate)
bp <- bp %>% mutate(time.from.koh1 = round(as.numeric(difftime(BPDate, KOH.start.dt, units="days"))))
bp <- bp %>% mutate(koh.cat = case_when(KOH.none == 1 ~ 'None',
                                        KOH.one == 1 ~ 'One',
                                        KOH.mult == 1 ~ 'Multiple'))
bp$koh.cat <- factor(bp$koh.cat, levels=c("None","One","Multiple"))
bp$male <- ifelse(bp$Sex == 'M', 1, 0)
bp$month.exact <- bp$time.from.koh1 / 30.4

ggplot(data= bp, 
       aes(x=month.exact, y=Systolic, group=UniqueIdentifier, colour=koh.cat))+
  geom_line() + 
  xlab("Time from first KOH meeting (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

summary_bp <- bp %>% group_by(koh.cat, month.exact) %>%
  summarise(mean_bp = mean(Systolic))

ggplot(data=summary_bp, 
       aes(x=month.exact, y=mean_bp, group=koh.cat, colour=koh.cat))+ 
   geom_smooth(method=loess, se=FALSE, linewidth=0.85) + geom_line(alpha=0.3) +
  xlab("Time from first KOH meeting (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance")) + scale_x_continuous(breaks=seq(0,12, by=3)) + scale_y_continuous(breaks=seq(65,215, by=25))
```
need to take out first bps for patients who don't have measurements within 1 year
taking out patients that don't have a post measurement within 1 year removes 7 patients: 6 none and 1 multiple. now there are 120 analyzable patients
by month: time x one: -1.45299597 (-3.81075425, 0.90476231) (p=0.23); time x multiple: -0.47911912 (-1.43399301,0.47575476) (p=0.32)
              one:    4.62355706 (-4.44647627, 13.69359039) (p=0.31); multiple: 2.44347920 (-4.79223538, 9.67919377) (p=0.50)
              time: 0.32698574 (-0.12857395, 0.78254542) (p=0.16)
              
For 1 year: time x one: -17.43595 (-45.66484, 10.79293); time x multiple: -5.749429 (-17.18191, 5.683051)
              one:  55.48268 (-52.18372, 163.1491); multiple: 29.32175 (-56.57026, 115.2138)
              time: 3.923829 (-1.53048, 9.378138)
```{r}
bp.mod1 <- lme(
  fixed = Systolic ~ month.exact * koh.cat + age + male + IncomeLevel + BLACERISK + avg.bmi,
  random = ~ month.exact | UniqueIdentifier,
  data = bp,
  method = "REML"
)
summary(bp.mod1)
intervals(bp.mod1)

fixed_effects <- summary(bp.mod1)$tTable

# Get estimate and SE for month.from.koh1
month_effect <- fixed_effects["month.exact", "Value"]
month_se <- fixed_effects["month.exact", "Std.Error"]

# Compute 12-month estimate and 95% CI
estimate_12m <- 12 * month_effect
se_12m <- 12 * month_se
ci_lower <- estimate_12m - 1.96 * se_12m
ci_upper <- estimate_12m + 1.96 * se_12m

# Display results
cat("12-month estimate:", estimate_12m, "\n")
cat("95% CI: (", ci_lower, ",", ci_upper, ")\n")

month_effect <- fixed_effects["koh.catOne", "Value"]
month_se <- fixed_effects["koh.catOne", "Std.Error"]

month_effect <- fixed_effects["koh.catMultiple", "Value"]
month_se <- fixed_effects["koh.catMultiple", "Std.Error"]

month_effect <- fixed_effects["month.exact:koh.catOne", "Value"]
month_se <- fixed_effects["month.exact:koh.catOne", "Std.Error"]

month_effect <- fixed_effects["month.exact:koh.catMultiple", "Value"]
month_se <- fixed_effects["month.exact:koh.catMultiple", "Std.Error"]
```


136 none; 27 one; 34 multiple; 197 total
```{r}
a1c <- read.csv('Analysis Data/Obj1A1c_LME.csv')[,-1]
length(unique(a1c$UniqueIdentifier))
```

removed 21 none, 1 one, and 2 multiple. now there are 197 analyzable patients
```{r cars}
# standardizing time as days from baseline measurement
a1c.first <- a1c %>% group_by(UniqueIdentifier) %>% arrange(A1cDate, .by_group=TRUE) %>% slice_head()
a1c.first$A1cDate <- a1c.first$KOH.start.dt
a1c <- a1c %>% filter(A1cDate>KOH.start.dt)

a1c <- rbind(a1c.first, a1c) %>% arrange(UniqueIdentifier, A1cDate)
a1c <- a1c %>% mutate(time.from.koh1 = round(as.numeric(difftime(A1cDate, KOH.start.dt, units="days"))))
a1c <- a1c %>% mutate(koh.cat = case_when(KOH.none == 1 ~ 'None',
                                        KOH.one == 1 ~ 'One',
                                        KOH.mult == 1 ~ 'Multiple'))
a1c$koh.cat <- factor(a1c$koh.cat, levels=c("None","One","Multiple"))
a1c$male <- ifelse(a1c$Sex == 'M', 1, 0)
a1c$month.exact <- a1c$time.from.koh1 / 30.4

ggplot(data= a1c, 
       aes(x=month.exact, y=A1c, group=UniqueIdentifier, colour=koh.cat))+
  geom_line() + 
  xlab("Time from first KOH meeting (months)") + ylab("A1c (%)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

summary_a1c <- a1c %>% group_by(koh.cat, month.exact) %>%
  summarise(mean_a1c = mean(A1c))

ggplot(data=summary_a1c, 
       aes(x=month.exact, y=mean_a1c, group=koh.cat, colour=koh.cat))+
   geom_smooth(method=loess, se=FALSE, linewidth=0.85) + geom_line(alpha=0.3) +
  xlab("Time from first KOH meeting (months)") + ylab("A1c (%)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance")) + scale_x_continuous(breaks=seq(0,12, by=3)) + scale_y_continuous(breaks=seq(5,15, by=2))

```

by month: time x one: -0.006648114 (-0.131476290,  0.118180063) (p=0.92); time x multiple: 0.023027198 (-0.070412100, 0.116466496) (p=0.63)
              one:  0.104844918 (-0.901726576, 1.111416412) (p=0.84); multiple: -0.410681954 (-1.319098984, 0.497735076) (p=0.37)
              time: -0.044443149 (-0.085665571, -0.003220728) (p=0.03)
              
For 1 year: time x one: -0.07977737 (-1.573278, 1.413723); time x multiple: 0.2763264 (-0.8416232,1.394276 )
              one:  1.258139 (-10.7436, 13.25987); multiple: -4.928183 (-15.75958, 5.903218)
              time: -0.5333178 (-1.026521, -0.0401143)
```{r}
a1c.mod1 <- lme(
  fixed = A1c ~ month.exact * koh.cat + age + male + IncomeLevel + BLACERISK + avg.bmi,
  random = ~ month.exact | UniqueIdentifier,
  data = a1c,
  method = "REML"
)
intervals(a1c.mod1)
summary(a1c.mod1)

fixed_effects <- summary(a1c.mod1)$tTable

# Get estimate and SE for month.from.koh1
month_effect <- fixed_effects["month.exact", "Value"]
month_se <- fixed_effects["month.exact", "Std.Error"]

# Compute 12-month estimate and 95% CI
estimate_12m <- 12 * month_effect
se_12m <- 12 * month_se
ci_lower <- estimate_12m - 1.96 * se_12m
ci_upper <- estimate_12m + 1.96 * se_12m

# Display results
cat("12-month estimate:", estimate_12m, "\n")
cat("95% CI: (",ci_lower,",", ci_upper,")\n")

month_effect <- fixed_effects["koh.catOne", "Value"]
month_se <- fixed_effects["koh.catOne", "Std.Error"]

month_effect <- fixed_effects["koh.catMultiple", "Value"]
month_se <- fixed_effects["koh.catMultiple", "Std.Error"]

month_effect <- fixed_effects["month.exact:koh.catOne", "Value"]
month_se <- fixed_effects["month.exact:koh.catOne", "Std.Error"]

month_effect <- fixed_effects["month.exact:koh.catMultiple", "Value"]
month_se <- fixed_effects["month.exact:koh.catMultiple", "Std.Error"]

```



secondary objective
126 marsh; 4695 NHW; 4821 total
```{r}
bp2 <- read.csv('Analysis Data/Obj2BP_LME.csv')[,-1]
length(unique(bp2$UniqueIdentifier[bp2$Marsh==1]))

bp2$KOHDate <- as.Date('2023-04-05')
bp2$Date <- as.Date(bp2$Date)
bp2.first <- bp2 %>% group_by(UniqueIdentifier) %>% arrange(Date, .by_group=TRUE) %>% slice_head()
bp2.first$Date <- bp2.first$KOHDate
bp2 <- bp2 %>% filter(Date>KOHDate)

bp2 <- rbind(bp2.first, bp2) %>% arrange(UniqueIdentifier, Date)
bp2 <- bp2 %>% mutate(time = round(as.numeric(difftime(Date, KOHDate, units="days"))))

bp2$male <- ifelse(bp2$Sex == 'M', 1, 0)
bp2$Marsh <- factor(bp2$Marsh)

bp2$month.exact <- bp2$time / 30.4
 
ggplot(data= bp2, 
       aes(x=month.exact, y=Systolic, group=UniqueIdentifier, colour=Marsh))+
  geom_line() + 
  xlab("Time from baseline (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() + 
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))

summary_bp2 <- bp2 %>% group_by(Marsh, month.exact) %>%
  summarise(mean_bp = mean(Systolic))

ggplot(data=summary_bp2, 
       aes(x=month.exact, y=mean_bp, group=Marsh, colour=Marsh))+
  geom_smooth(method=loess, se=FALSE, linewidth=0.85) + geom_line(alpha=0.3) +
  xlab("Time from baseline (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese")) + scale_x_continuous(breaks=seq(0,21, by=3)) + scale_y_continuous(breaks=seq(85,210, by=25))
```
by month: time x marsh: -6.199612e-02 (-2.248453e-01, 1.008531e-01) (p=0.46); marsh: 1.434512 (-9.570280e-01, 3.826051e+00) (p=0.24);
              time: -4.298336e-02 (-6.934832e-02, -1.661841e-02) (p=0.0014)

For 1 year: time x marsh: -0.7439535 (-2.698112, 1.210205); marsh: 17.21414 (-11.47765, 45.90593);
              time: -0.5158004 (-0.8321747, -0.199426)
```{r}
bp.mod2 <- lme(
  fixed = Systolic ~ month.exact * Marsh + age + male + IncomeLevel + avg.bmi,
  random = ~ month.exact | UniqueIdentifier,
  data = bp2,
  method = "REML"
)
summary(bp.mod2)
intervals(bp.mod2)

fixed_effects <- summary(bp.mod2)$tTable

# Get estimate and SE for month.from.koh1
month_effect <- fixed_effects["month.exact", "Value"]
month_se <- fixed_effects["month.exact", "Std.Error"]

# Compute 12-month estimate and 95% CI
estimate_12m <- 12 * month_effect
se_12m <- 12 * month_se
ci_lower <- estimate_12m - 1.96 * se_12m
ci_upper <- estimate_12m + 1.96 * se_12m

# Display results
cat("12-month estimate:", estimate_12m, "\n")
cat("95% CI: (", ci_lower, ",", ci_upper, ")\n")

month_effect <- fixed_effects["Marsh1", "Value"]
month_se <- fixed_effects["Marsh1", "Std.Error"]

month_effect <- fixed_effects["month.exact:Marsh1", "Value"]
month_se <- fixed_effects["month.exact:Marsh1", "Std.Error"]

```

objective 2 A1c
210 marsh; 2349 NHW; 2559 total
```{r}
a1c2 <- read.csv('Analysis Data/Obj2A1c_LME.csv')[,-1]
length(unique(a1c2$UniqueIdentifier))

a1c2$KOHDate <- as.Date('2023-04-05')
a1c2$Date <- as.Date(a1c2$Date)
a1c2.first <- a1c2 %>% group_by(UniqueIdentifier) %>% arrange(Date, .by_group=TRUE) %>% slice_head()
a1c2.first$Date <- a1c2.first$KOHDate
a1c2 <- a1c2 %>% filter(Date>KOHDate)

a1c2 <- rbind(a1c2.first, a1c2) %>% arrange(UniqueIdentifier, Date)
a1c2 <- a1c2 %>% mutate(time = round(as.numeric(difftime(Date, KOHDate, units="days"))))

a1c2$male <- ifelse(a1c2$Sex == 'M', 1, 0)
a1c2$Marsh <- factor(a1c2$Marsh)
a1c2$month.exact <- a1c2$time / 30.4

ggplot(data= a1c2, 
       aes(x=month.exact, y=A1c, group=UniqueIdentifier, colour=Marsh))+
  geom_line() +
  xlab("Time from baseline (months)") + ylab("A1c (%)") + theme_bw() + 
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))

summary_a1c2 <- a1c2 %>% group_by(Marsh, month.exact) %>%
  summarise(mean_a1c = mean(A1c))

ggplot(data=summary_a1c2, 
       aes(x=month.exact, y=mean_a1c, group=Marsh, colour=Marsh))+
  geom_smooth(method=loess, se=FALSE, linewidth=0.85) + geom_line(alpha=0.3) +
  xlab("Time from baseline (months)") + ylab("A1c (%)") + theme_bw() + 
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese")) + scale_x_continuous(breaks=seq(0,21, by=3)) + scale_y_continuous(breaks=seq(4.5, 17, by=2.5))
```
by month: time x marsh: -1.081642e-02 (-0.024975889, 3.343046e-03) (p=0.13); marsh: 2.240465 (1.988022603, 2.492907) (p=0.000);
              time: 1.109435e-03 (-0.002975881, 5.194750e-03) (p=0.59)

For 1 year: time x marsh: -0.1297971 (-0.2996862, 0.04009212); marsh: 26.88558 (23.85765, 29.9135);
              time: 0.01331322 (-0.03570352, 0.06232995)
```{r}
a1c.mod2 <- lme(
  fixed = A1c ~ month.exact * Marsh + age + male + IncomeLevel + avg.bmi,
  random = ~ month.exact | UniqueIdentifier,
  data = a1c2,
  method = "REML"
)
summary(a1c.mod2)
intervals(a1c.mod2)

fixed_effects <- summary(a1c.mod2)$tTable

# Get estimate and SE for month.from.koh1
month_effect <- fixed_effects["month.exact", "Value"]
month_se <- fixed_effects["month.exact", "Std.Error"]

# Compute 12-month estimate and 95% CI
estimate_12m <- 12 * month_effect
se_12m <- 12 * month_se
ci_lower <- estimate_12m - 1.96 * se_12m
ci_upper <- estimate_12m + 1.96 * se_12m

# Display results
cat("12-month estimate:", estimate_12m, "\n")
cat("95% CI: (", ci_lower, ",", ci_upper, ")\n")

month_effect <- fixed_effects["Marsh1", "Value"]
month_se <- fixed_effects["Marsh1", "Std.Error"]

month_effect <- fixed_effects["month.exact:Marsh1", "Value"]
month_se <- fixed_effects["month.exact:Marsh1", "Std.Error"]
```

```{r}
options(digits=3)
# obj 1 SBP regression result table
###########################################
# format for regression results from models
###########################################

coef_names = c("month.exact:koh.catOne", "month.exact:koh.catMultiple", "koh.catOne", "koh.catMultiple", "month.exact")
col_names = c("est.", "lower", "upper")

# primary objective
obj1_res = data.frame(
  rbind(
    cbind(intervals(a1c.mod1)$fixed[coef_names, col_names], summary(a1c.mod1)$tTable[coef_names, "p-value"]),
    cbind(intervals(bp.mod1, which="fixed")$fixed[coef_names, col_names], summary(bp.mod1)$tTable[coef_names, "p-value"])
  )
)

# obj1_res = data.frame(
#   est = c(summary(mod_o1_a1c)$tTable[coef_names, "Value"], summary(mod_o1_bp)$tTable[coef_names, "Value"]),
#   CI = intervals(mod_o1_a1c)$fixed[coef_names, col_names]
# )

rownames(obj1_res) = c("months:KOHOnce", "months:KOHMultiple", "KOHOnce", "KOHMultiple", "months", "months:KOHOnce ", "months:KOHMultiple ", "KOHOnce ", "KOHMultiple ", "months ")
colnames(obj1_res) = c("estimate", "95% CI lower", "95% CI upper", "p-value")

kable(obj1_res) %>% pack_rows("A1c longitudinal", 1, 5) %>% pack_rows("SBP longitudinal", 6, 10)

# secondary objective
coef_names = c("month.exact:Marsh1", "Marsh1", "month.exact")
col_names = c("est.", "lower", "upper")

obj2_res = data.frame(
  rbind(
    cbind(intervals(a1c.mod2, which="fixed")$fixed[coef_names, col_names], summary(a1c.mod2)$tTable[coef_names, "p-value"]),
    cbind(intervals(bp.mod2, which="fixed")$fixed[coef_names, col_names], summary(bp.mod2)$tTable[coef_names, "p-value"])
  )
)

# obj1_res = data.frame(
#   est = c(summary(mod_o1_a1c)$tTable[coef_names, "Value"], summary(mod_o1_bp)$tTable[coef_names, "Value"]),
#   CI = intervals(mod_o1_a1c)$fixed[coef_names, col_names]
# )

rownames(obj2_res) = c("months:Marshallese","Marshallese", "months", "months:Marshallese ","Marshallese ", "months ")
colnames(obj2_res) = c("estimate", "95% CI lower", "95% CI upper", "p-value")

kable(obj2_res) %>% pack_rows("A1c longitudinal", 1, 3) %>% pack_rows("SBP longitudinal", 4, 6)
```
