---
title: "Objective 1 Longitudinal Analysis"
author: "Gabby Lopez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)
```
note to self: run regression diagnostics and ignore clustering - regular linear regression and see if any observations are having huge impact
change days to months

```{r}
bp <- read.csv('Analysis Data/Obj1BP_LME.csv')[,-1]
```

```{r cars}
# standardizing time as days from baseline measurement
bp.first <- bp %>% group_by(UniqueIdentifier) %>% arrange(BPDate, .by_group=TRUE) %>% slice_head()
bp.first$BPDate <- bp.first$KOH.start.dt
bp <- bp %>% filter(BPDate>KOH.start.dt)

bp <- rbind(bp.first, bp) %>% arrange(UniqueIdentifier, BPDate)
bp <- bp %>% mutate(time.from.koh1 = round(as.numeric(difftime(BPDate, KOH.start.dt, units="days"))))
bp <- bp %>% mutate(koh.cat = case_when(KOH.none == 1 ~ 'None',
                                        KOH.one == 1 ~ 'One',
                                        KOH.mult == 1 ~ 'Multiple'))
bp$koh.cat <- factor(bp$koh.cat, levels=c("None","One","Multiple"))
bp$male <- ifelse(bp$Sex == 'M', 1, 0)
bp$month.exact <- bp$time.from.koh1 / 30.4
bp <- bp %>% mutate(month.from.koh1 = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 1 ~ 1,
                                                month.exact > 1 & month.exact <= 2 ~ 2,
                                                month.exact > 2 & month.exact <= 3 ~ 3,
                                                month.exact > 3 & month.exact <= 4 ~ 4,
                                                month.exact > 4 & month.exact <= 5 ~ 5,
                                                month.exact > 5 & month.exact <= 6 ~ 6,
                                                month.exact > 6 & month.exact <= 7 ~ 7,
                                                month.exact > 7 & month.exact <= 8 ~ 8,
                                                month.exact > 8 & month.exact <= 9 ~ 9,
                                                month.exact > 9 & month.exact <= 10 ~ 10,
                                                month.exact > 10 & month.exact <= 11 ~ 11,
                                                month.exact > 11 & time.from.koh1 <= 365 ~ 12))

bp <- bp %>% mutate(quart.from.koh1 = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 3 ~ 1,
                                                month.exact > 3 & month.exact <= 6 ~ 2,
                                                month.exact > 6 & month.exact <= 9 ~ 3,
                                                month.exact > 9 & time.from.koh1 <= 365 ~ 4))

ggplot(data= bp, 
       aes(x=month.from.koh1, y=Systolic, group=UniqueIdentifier, colour=koh.cat))+
  geom_line() + 
  xlab("Time (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

ggplot(data= bp, 
       aes(x=quart.from.koh1, y=Systolic, group=UniqueIdentifier, colour=koh.cat))+
  geom_line() +
  xlab("Time (3 months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

summary_bp <- bp %>% group_by(koh.cat, month.from.koh1) %>%
  summarise(mean_bp = mean(Systolic))

ggplot(data=summary_bp, 
       aes(x=month.from.koh1, y=mean_bp, group=koh.cat, colour=koh.cat))+ 
   geom_smooth(method=loess, se=FALSE, linewidth=0.75) + 
  xlab("Time (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

summary_bp <- bp %>% group_by(koh.cat, quart.from.koh1) %>%
  summarise(mean_bp = mean(Systolic))

ggplot(data=summary_bp, 
       aes(x=quart.from.koh1, y=mean_bp, group=koh.cat, colour=koh.cat))+ 
   geom_smooth(method=loess, se=FALSE, linewidth=0.75) + 
  xlab("Time (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))
```
need to take out first bps for patients who don't have measurements within 1 year
taking out patients that don't have a post measurement within 1 year removes 7 patients: 6 none and 1 multiple. now there are 120 analyzable patients
by month: time x one: 	-1.52430841 (-3.75803035, 0.70941354) (p=0.18); time x multiple: -0.49815691 (-1.41733394, 0.42102011) (p=0.29)
              one:    5.27758690 (-4.03307467, 14.58824848) (p=0.26); multiple: 2.64455146 (-4.68902241 , 9.97812532) (p=0.48)
              time: 0.32826116 (-0.11001054 , 0.76653286) (p=0.14)
              
by quarter: time x one: 	-2.28608584 (-8.36103190, 3.78886021) (p=0.46); time x multiple: -1.40703231 (-4.07497335, 1.26090873) (p=0.30)
              one:    4.32897744 (-5.56940096, 14.22735584) (p=0.39); multiple: 2.87344149 (-4.76596121, 10.51284419) (p=0.46)
              time: 1.01610547 (-0.25623046, 2.28844140) (p=0.12)
```{r}
bp.mod1 <- lme(
  fixed = Systolic ~ month.from.koh1 * koh.cat + age + male + IncomeLevel + BLACERISK + avg.bmi,
  random = ~ month.from.koh1 | UniqueIdentifier,
  data = bp,
  method = "REML"
)
summary(bp.mod1)
intervals(bp.mod1)

bp.mod1.2 <- lme(
  fixed = Systolic ~ quart.from.koh1 * koh.cat + age + male + IncomeLevel + BLACERISK + avg.bmi,
  random = ~ quart.from.koh1 | UniqueIdentifier,
  data = bp,
  method = "REML"
)
summary(bp.mod1.2)
intervals(bp.mod1.2)
```

```{r}
a1c <- read.csv('Analysis Data/Obj1A1c_LME.csv')[,-1]
```

removed 21 none, 1 one, and 2 multiple. now there are 197 analyzable patients
```{r cars}
# standardizing time as days from baseline measurement
a1c.first <- a1c %>% group_by(UniqueIdentifier) %>% arrange(A1cDate, .by_group=TRUE) %>% slice_head()
a1c.first$A1cDate <- a1c.first$KOH.start.dt
a1c <- a1c %>% filter(A1cDate>KOH.start.dt)

a1c <- rbind(a1c.first, a1c) %>% arrange(UniqueIdentifier, A1cDate)
a1c <- a1c %>% mutate(time.from.koh1 = round(as.numeric(difftime(A1cDate, KOH.start.dt, units="days"))))
a1c <- a1c %>% mutate(koh.cat = case_when(KOH.none == 1 ~ 'None',
                                        KOH.one == 1 ~ 'One',
                                        KOH.mult == 1 ~ 'Multiple'))
a1c$koh.cat <- factor(a1c$koh.cat, levels=c("None","One","Multiple"))
a1c$male <- ifelse(a1c$Sex == 'M', 1, 0)
a1c$month.exact <- a1c$time.from.koh1 / 30.4
a1c <- a1c %>% mutate(month.from.koh1 = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 1 ~ 1,
                                                month.exact > 1 & month.exact <= 2 ~ 2,
                                                month.exact > 2 & month.exact <= 3 ~ 3,
                                                month.exact > 3 & month.exact <= 4 ~ 4,
                                                month.exact > 4 & month.exact <= 5 ~ 5,
                                                month.exact > 5 & month.exact <= 6 ~ 6,
                                                month.exact > 6 & month.exact <= 7 ~ 7,
                                                month.exact > 7 & month.exact <= 8 ~ 8,
                                                month.exact > 8 & month.exact <= 9 ~ 9,
                                                month.exact > 9 & month.exact <= 10 ~ 10,
                                                month.exact > 10 & month.exact <= 11 ~ 11,
                                                month.exact > 11 & time.from.koh1 <= 365 ~ 12))

a1c <- a1c %>% mutate(quart.from.koh1 = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 3 ~ 1,
                                                month.exact > 3 & month.exact <= 6 ~ 2,
                                                month.exact > 6 & month.exact <= 9 ~ 3,
                                                month.exact > 9 & time.from.koh1 <= 365 ~ 4))

ggplot(data= a1c, 
       aes(x=month.from.koh1, y=A1c, group=UniqueIdentifier, colour=koh.cat))+
  geom_line() + 
  xlab("Time (months)") + ylab("A1c") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

summary_a1c <- a1c %>% group_by(koh.cat, month.from.koh1) %>%
  summarise(mean_a1c = mean(A1c))
ggplot(data=summary_a1c, 
       aes(x=month.from.koh1, y=mean_a1c, group=koh.cat, colour=koh.cat))+
   geom_smooth(method=loess, se=FALSE, linewidth=0.75) + 
  xlab("Time (months)") + ylab("A1c (%)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

summary_a1c <- a1c %>% group_by(koh.cat, quart.from.koh1) %>%
  summarise(mean_a1c = mean(A1c))

ggplot(data=summary_a1c, 
       aes(x=quart.from.koh1, y=mean_a1c, group=koh.cat, colour=koh.cat))+
   geom_smooth(method=loess, se=FALSE, linewidth=0.75) + 
  xlab("Time (3 months)") + ylab("A1c (%)") + theme_bw() +
  guides(colour = guide_legend(title = "KOH Attendance"))

```

by month: time x one: 	-0.014026759 (-0.133298463, 0.105244946) (p=0.82); time x multiple: 0.022737160 (-0.067026719, 0.112501039) (p=0.62)
              one:    0.126892650 (-0.888169071, 1.141954370) (p=0.81); multiple: -0.415746454 (-1.331031485, 0.499538577) (p=0.37)
              time: -0.043508976 (-0.083147648, -0.003870304) (p=0.03)
by quarter: time x one: 	-0.082208809 (-0.411858937, 0.247441318) (p=0.62); time x multiple: 0.080900676 (-0.173814468, 0.335615820) (p=0.53)
              one:    0.215292969 (-0.816583452, 1.247169391) (p=0.68); multiple: -0.452110468 (-1.378131399, 0.473910464) (p=0.34)
              time: -0.123239343 (-0.234286484, -0.012192201) (p=0.03)
```{r}
a1c.mod1 <- lme(
  fixed = A1c ~ month.from.koh1 * koh.cat + age + male + IncomeLevel + BLACERISK + avg.bmi,
  random = ~ month.from.koh1 | UniqueIdentifier,
  data = a1c,
  method = "REML"
)
intervals(a1c.mod1)
summary(a1c.mod1)

a1c.mod1.2 <- lme(
  fixed = A1c ~ quart.from.koh1 * koh.cat + age + male + IncomeLevel + BLACERISK + avg.bmi,
  random = ~ quart.from.koh1 | UniqueIdentifier,
  data = a1c,
  method = "REML"
)
intervals(a1c.mod1.2)
summary(a1c.mod1.2)
```



secondary objective

```{r}
bp2 <- read.csv('Analysis Data/Obj2BP_LME.csv')[,-1]

bp2$KOHDate <- as.Date('2023-04-05')
bp2$Date <- as.Date(bp2$Date)
bp2.first <- bp2 %>% group_by(UniqueIdentifier) %>% arrange(Date, .by_group=TRUE) %>% slice_head()
bp2.first$Date <- bp2.first$KOHDate
bp2 <- bp2 %>% filter(Date>KOHDate)

bp2 <- rbind(bp2.first, bp2) %>% arrange(UniqueIdentifier, Date)
bp2 <- bp2 %>% mutate(time = round(as.numeric(difftime(Date, KOHDate, units="days"))))

bp2$male <- ifelse(bp2$Sex == 'M', 1, 0)
bp2$Marsh <- factor(bp2$Marsh)

bp2$month.exact <- bp2$time / 30.4
bp2 <- bp2 %>% mutate(month = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 1 ~ 1,
                                                month.exact > 1 & month.exact <= 2 ~ 2,
                                                month.exact > 2 & month.exact <= 3 ~ 3,
                                                month.exact > 3 & month.exact <= 4 ~ 4,
                                                month.exact > 4 & month.exact <= 5 ~ 5,
                                                month.exact > 5 & month.exact <= 6 ~ 6,
                                                month.exact > 6 & month.exact <= 7 ~ 7,
                                                month.exact > 7 & month.exact <= 8 ~ 8,
                                                month.exact > 8 & month.exact <= 9 ~ 9,
                                                month.exact > 9 & month.exact <= 10 ~ 10,
                                                month.exact > 10 & month.exact <= 11 ~ 11,
                                                month.exact > 11 & month.exact <= 12 ~ 12,
                                                month.exact > 12 & month.exact <= 13 ~ 13,
                                                month.exact > 13 & month.exact <= 14 ~ 14,
                                                month.exact > 14 & month.exact <= 15 ~ 15,
                                                month.exact > 15 & month.exact <= 16 ~ 16,
                                                month.exact > 16 & month.exact <= 17 ~ 17,
                                                month.exact > 17 & month.exact <= 18 ~ 18,
                                                month.exact > 18 & month.exact <= 19 ~ 19,
                                                month.exact > 19 & month.exact <= 20 ~ 20,
                                                month.exact > 20 & month.exact <= 21 ~ 21,
                                                month.exact > 21 & month.exact <= 22 ~ 22))
bp2 <- bp2 %>% mutate(quarter = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 3 ~ 1,
                                                month.exact > 3 & month.exact <= 6 ~ 2,
                                                month.exact > 6 & month.exact <= 9 ~ 3,
                                                month.exact > 9 & month.exact <= 12 ~ 4,
                                                month.exact > 12 & month.exact <= 15 ~ 5,
                                                month.exact > 15 & month.exact <= 18 ~ 6,
                                                month.exact > 18 & month.exact <= 21 ~ 7,
                                                month.exact > 21 & month.exact <= 24 ~ 8))

ggplot(data= bp2, 
       aes(x=month, y=Systolic, group=UniqueIdentifier, colour=Marsh))+
  geom_line() + 
  xlab("Time (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() + 
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))

summary_bp2 <- bp2 %>% group_by(Marsh, month) %>%
  summarise(mean_bp = mean(Systolic))

ggplot(data=summary_bp2, 
       aes(x=month, y=mean_bp, group=Marsh, colour=Marsh))+
  geom_smooth(method=loess, se=FALSE) + 
  xlab("Time (months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))

summary_bp2 <- bp2 %>% group_by(Marsh, quarter) %>%
  summarise(mean_bp = mean(Systolic))

ggplot(data=summary_bp2, 
       aes(x=quarter, y=mean_bp, group=Marsh, colour=Marsh))+
  geom_smooth(method=loess, se=FALSE) + 
  xlab("Time (3 months)") + ylab("Systolic Blood Pressure (mmHg)") + theme_bw() +
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))
```
by month: time x marsh: 	9.967987e-04 (-1.650823e-01, 1.630887e-01) (p=0.99); marsh: 1.151615e+00 (-1.275721e+00, 3.578951e+00) (p=0.35);
              time: -4.973476e-02 (-7.628624e-02, -2.318328e-02) (p=0.0002)
by quarter: time x marsh: 	-2.604872e-03 (-4.791694e-01, 4.739596e-01) (p=0.99); marsh: 1.172658e+00 (1.303669e+00, 3.648985e+00) (p=0.35);
              time: -1.472664e-01 (-2.245122e-01, -7.002047e-02) (p=0.0002)
```{r}
bp.mod2 <- lme(
  fixed = Systolic ~ month * Marsh + age + male + IncomeLevel + avg.bmi,
  random = ~ month | UniqueIdentifier,
  data = bp2,
  method = "REML"
)
summary(bp.mod2)
intervals(bp.mod2)

bp.mod2.2 <- lme(
  fixed = Systolic ~ quarter * Marsh + age + male + IncomeLevel + avg.bmi,
  random = ~ quarter | UniqueIdentifier,
  data = bp2,
  method = "REML"
)
summary(bp.mod2.2)
intervals(bp.mod2.2)
```

objective 2 A1c
```{r}
a1c2 <- read.csv('Analysis Data/Obj2A1c_LME.csv')[,-1]

a1c2$KOHDate <- as.Date('2023-04-05')
a1c2$Date <- as.Date(a1c2$Date)
a1c2.first <- a1c2 %>% group_by(UniqueIdentifier) %>% arrange(Date, .by_group=TRUE) %>% slice_head()
a1c2.first$Date <- a1c2.first$KOHDate
a1c2 <- a1c2 %>% filter(Date>KOHDate)

a1c2 <- rbind(a1c2.first, a1c2) %>% arrange(UniqueIdentifier, Date)
a1c2 <- a1c2 %>% mutate(time = round(as.numeric(difftime(Date, KOHDate, units="days"))))

a1c2$male <- ifelse(a1c2$Sex == 'M', 1, 0)
a1c2$Marsh <- factor(a1c2$Marsh)
a1c2$month.exact <- a1c2$time / 30.4
a1c2 <- a1c2 %>% mutate(month = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 1 ~ 1,
                                                month.exact > 1 & month.exact <= 2 ~ 2,
                                                month.exact > 2 & month.exact <= 3 ~ 3,
                                                month.exact > 3 & month.exact <= 4 ~ 4,
                                                month.exact > 4 & month.exact <= 5 ~ 5,
                                                month.exact > 5 & month.exact <= 6 ~ 6,
                                                month.exact > 6 & month.exact <= 7 ~ 7,
                                                month.exact > 7 & month.exact <= 8 ~ 8,
                                                month.exact > 8 & month.exact <= 9 ~ 9,
                                                month.exact > 9 & month.exact <= 10 ~ 10,
                                                month.exact > 10 & month.exact <= 11 ~ 11,
                                                month.exact > 11 & month.exact <= 12 ~ 12,
                                                month.exact > 12 & month.exact <= 13 ~ 13,
                                                month.exact > 13 & month.exact <= 14 ~ 14,
                                                month.exact > 14 & month.exact <= 15 ~ 15,
                                                month.exact > 15 & month.exact <= 16 ~ 16,
                                                month.exact > 16 & month.exact <= 17 ~ 17,
                                                month.exact > 17 & month.exact <= 18 ~ 18,
                                                month.exact > 18 & month.exact <= 19 ~ 19,
                                                month.exact > 19 & month.exact <= 20 ~ 20,
                                                month.exact > 20 & month.exact <= 21 ~ 21,
                                                month.exact > 21 & month.exact <= 22 ~ 22))
a1c2 <- a1c2 %>% mutate(quarter = case_when(month.exact == 0 ~ 0,
                                                month.exact > 0 & month.exact <= 3 ~ 1,
                                                month.exact > 3 & month.exact <= 6 ~ 2,
                                                month.exact > 6 & month.exact <= 9 ~ 3,
                                                month.exact > 9 & month.exact <= 12 ~ 4,
                                                month.exact > 12 & month.exact <= 15 ~ 5,
                                                month.exact > 15 & month.exact <= 18 ~ 6,
                                                month.exact > 18 & month.exact <= 21 ~ 7,
                                                month.exact > 21 & month.exact <= 24 ~ 8))

ggplot(data= a1c2, 
       aes(x=month, y=A1c, group=UniqueIdentifier, colour=Marsh))+
  geom_line() +
  xlab("Time (months)") + ylab("A1c") + theme_bw() + 
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))

summary_a1c2 <- a1c2 %>% group_by(Marsh, month) %>%
  summarise(mean_a1c = mean(A1c))

ggplot(data=summary_a1c2, 
       aes(x=month, y=mean_a1c, group=Marsh, colour=Marsh))+
  geom_smooth(method=loess, se=FALSE) +
  xlab("Time (months)") + ylab("A1c (%)") + theme_bw() + 
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))

summary_a1c2 <- a1c2 %>% group_by(Marsh, quarter) %>%
  summarise(mean_a1c = mean(A1c))

ggplot(data=summary_a1c2, 
       aes(x=quarter, y=mean_a1c, group=Marsh, colour=Marsh))+
  geom_smooth(method=loess, se=FALSE) +
  xlab("Time (3 months)") + ylab("A1c (%)") + theme_bw() + 
  scale_color_discrete(name = "Race", labels=c("Non-Hispanic White", "Marshallese"))
```
by month: time x marsh: 	-1.089352e-02 (-0.0247203817, 2.933351e-03) (p=0.12); marsh: 2.245293e+00 (1.9917854763, 2.498800e+00) (p=0.000);
              time: 1.125405e-03 (-0.0028656002, 5.116410e-03) (p=0.58)
by quarter: time x marsh: 	-3.029897e-02 (-0.0698752134, -1.402562e-02) (p=0.13); marsh:  2.245818e+00 (1.9899852173, 2.501651e+00) (p=0.000);
              time: 3.013347e-03 (-0.0084541148, 1.448081e-02) (p=0.61)
```{r}
a1c.mod2 <- lme(
  fixed = A1c ~ month * Marsh + age + male + IncomeLevel + avg.bmi,
  random = ~ month | UniqueIdentifier,
  data = a1c2,
  method = "REML"
)
summary(a1c.mod2)
intervals(a1c.mod2)

a1c.mod2.2 <- lme(
  fixed = A1c ~ quarter * Marsh + age + male + IncomeLevel + avg.bmi,
  random = ~ quarter | UniqueIdentifier,
  data = a1c2,
  method = "REML"
)
summary(a1c.mod2.2)
intervals(a1c.mod2.2)
```

